{Simulating the Brownian Dynamics of a Two-dimensional Model for the Dynein Motor Protein}\vspace{1.5em}\\ \includegraphics[width=0.5\columnwidth]{OSU_logo.png}\vspace{0.5em}\\ \large{Oregon State University}

### 2019-03-20 papers/waczak-thesis/thesis/chapters/dynein.tex
big thesis update

\section{Directed Motion}
\section{Drunken Walking}
\section{Transporting Cargo}
\section{Experimental Stepping Statistics}

>>>>

\section{The microtubule}
We've already mentioned that cells employ motor proteins and the cytoskeleton to solve the cellular transport problem. The following figure demonstrates the complexity of this network of cellular highways within endothelial cells. 
\begin{figure}[!hbt]
	\centering
	\includegraphics[width=0.5\columnwidth]{cytoskeleton}
	\caption{Endothelial cells with their cytoskeleton structures shown. The nucleus of each cell is colored in blue. Actin filaments are shown in red and tubulin is shown in green. Image taken from \cite{cytoskeleton}}
	\label{fig:cytoskeleton}
\end{figure}

Dynein walks along a particular cyctoskeleton filament called a microtubule. Microtubules are cytoskeletal structures formed by self assembly of $\alpha$ and $\beta$ tubulin heterodimers (shown as green in Figure \ref{fig:cytoskeleton}). \cite{downing_tubulin_1998} These molecular structures are typically comprised of 13 protofilaments each of which is a periodic arrangement of the $\alpha$ and $\beta$ dimers.

Microtubules exhibit an inherent polarity identified as the plus ($+$) and minus ($-$) ends \cite{lodish_microtubule_2000}. This polarity is both function and structural. One end of the microtubule experiences much faster addition of tubulin making it easy to identify\cite{heidemann_visualization_1980}. For our purposes though, the structural polarity of the microtubule is more important as the formation of the microtubule results in a helical structure that has definite chirality \cite{microtubule_chirality}. Two possible microtubule configurations are shown in Figure \ref{fig:chirality}. Note that even though there is a deviation by angle $\theta$ between the two microtubules, the overall handedness of the helix is the same.

\begin{figure}[!hbt]
	\centering 
	\includegraphics[width=.5\columnwidth]{microtubule_chirality}
	\caption{Two possible microtubule structures illustrating the chirality of the helix as well as the skew angle of the dimers of which only one is shown for illustrative purposes. Image taken from \cite{microtubule_chirality}.}
	\label{fig:chirality}
\end{figure}

As Figure \ref{fig:cytoskeleton} shows, microtubules foliate the entire cell with paths that enable motor proteins to move cargo. Most(not all) variations of the kinesin molecular motor are plus-end directed. They are almost always observed to walk towards the plus-end of the microtubule. Dynein is strictly a minus-end directed motor \cite{alberts_molecular_2002}. How this directed motion is derived from the symmetric dynein structure is still an open question.\\

\section{The powerstroke model} 
Take another look at Figure \ref{fig:motor proteins}. How does this complicated structure perform complex locomotion using only the resources of its aqueous environment? How can a highly symmetric structure lead to directed motion? The chemical resolution to this problem arises from the interaction of ATP with each of the motor domains of dynein. The cycle of converting the chemical energy of  ATP into useful mechanical energy for motion is called the \textit{mechanochemical cycle}\cite{cianfrocco_mechanism_2015}. Describing all of the chemical interactions involved in a single dynein step has been an active area of research. The currently accepted model was proposed by Cianfrocco et al. and is summarized in the following graphic shown in Figure \ref{fig:mechanochemical_cycle}.

\begin{figure}[!hbt]
	\centering
	\includegraphics[width=1\columnwidth]{mechanochemical_cycle}
	\caption{The mechanochemical cycle of dynein. Dynein goes through many identifiable stages as ATP from the environment interacts with the motor domain and changes the conformation of the protein. Image taken from \cite{cianfrocco_mechanism_2015}.}
	\label{fig:mechanochemical_cycle}
\end{figure}

The important takeaway from Figure \ref{fig:mechanochemical_cycle} is that we can identify a ``pre-stroke" and a ``post-stroke" state corresponding to the conformational change that occurs in the purple linker between steps 7 and 8. These states are again illustrated in the following figure to make the physical change in shape clear.

\begin{figure}[!hbt]
	\centering
	\includegraphics[width=1\columnwidth]{pre_post_powerstroke}
	\caption{An illustration of the pre and post stroke states of dynein in relation to the microtubule and the cargo that they carry. This two-dimensional illustration shows only one of legs of dynein. Image taken from \cite{dynein_powerstroke}.}
\end{figure}

Scientists have determined the steps of this chemical cycle by carefully isolating and analyzing the resulting structures via cryo-electron microscopy \cite{cryo_em}. While this is useful for determining the conformation before and after a chemical process occurs, it does note describe the actual motion along the microtubule-- only intermittent states. All of the information about how flexible each domain is during binding and to what extent the protein diffuses in its search for the microtubule require alternative methods to establish.

\section{Other observational techniques} 
 Kinesin and Dynein have masses of a few hundred kDa and  $\sim2$ MDa respectively \cite{liao_kinesin_1998, johnson_structure_1983}. Recalling that 1 Dalton (Da) corresponds to one atomic mass unit means these proteins consist of, at most, a few million atoms. This small scale poses a unique experimental challenge. One way this is approached at Prof. Weihong Qiu's group is by attaching small fluorescent molecules called fluorophores to the binding sites that motor proteins use to latch onto their cargo. Then, the individual movement of motors can be tracked using total internal reflection fluorescence microscopy to produce video of the proteins in action as illustrated in figure \ref{fig:weihong tirf} from \cite{qiu_dynein_2012}.

\begin{figure}[!hbt]
	\centering
	\includegraphics[angle=90, width=0.75\columnwidth]{weihong_motor}
	\caption{\textbf{TIRF Microscopy Imaging} individual kinesin proteins are imaged as small green dots and tracked as they move. Slices at different times are stacked so that the slope of each green line indicates the velocity of the protein. Image taken from \cite{qiu_dynein_2012}.}
	\label{fig:weihong tirf}
\end{figure}

Figure \ref{fig:weihong tirf} demonstrates fluorescence microscopy enables observation of protein speed, how far it walks along the microtubule, which direction is preferred. Furthermore, by tagging a motor domain instead of the tail, this technique can resolve step sizes and as shown in Figure \ref{fig:weihong histogram}. \cite{qiu_dynein_2012} Taking into account these studies as well as the delicate cryo-em analysis forms a rough sketch for dynein's motion. Unfortunately though, this is the limit of current experimental methods; observing the individual motions of each of the binding domains of a dynein protein is not yet possible.

\todo[inline]{TODO: Add in discussion of using gold particles + scattering to observe Brownian wiggling. This was a result of conversations with Colloquium speaker Kazuhiro Oiwa.}

\section{Experimental stepping statistics}
Here I will include the various stepping statistics (distribution, binding/unbinding rates, etc...) that we hope to fit our simulation to.

\section{More motivation} 
Here I will discuss other simulation efforts that, in our opinion, make far to many assumptions or are purely monte carlo and don't actually simulate the motion of the protein. I will also discuss why our simulation is a happy medium between a full on molecular dynamics simulation (time expensive but accurate) and monte carlo (very fast but doesn't dynamically simulate the protein and instead just selects steps from a distribution).

### 2019-03-20 papers/waczak-thesis/thesis/chapters/introduction.tex
big thesis update

Cells are the fundamental unit for all life on Earth. In order
for them to perform their biological tasks, the complex structures they contain
must be able to replicate and move within their aqueous environment. This
transport problem  poses a unique challenge as diffusion alone is far to slow,
and random, a process to be able to produce movement on cellular timescales[].
Nature's solution to this problem is twofold. Within each cell there is a vast network of interweaving filaments called the \textit{cytoskeleton} that spans the gaps between the substructures of the cell. This cytoskeleton serves as a highway to a class of proteins called \textit{molecular motors} which can be distinguished into three types: kinesin, myosin, and dynein. Molecular motors attach to cellular cargo and drag it along the cytoskeleton by hydrolyzing Adenosine triphosphate (ATP) in order to convert stored chemical energy into mechanical work\cite{alberts_molecular_2002}. Correct operation of these motors results in processes such as cell division, muscle contraction, and axonal transport; failure can lead to debilitating diseases such as ALS \cite{boillee_als:_2006}.

@@

Astonishingly, each of these motors is composed of two legs that combine to form a bipedal structure (see figure \ref{fig:motor proteins}). that steps like a human as it moves along the cytoskeleton. Myosin and kinesin behave like a normal person; their steps alternate between opposing domains. This motion is called \textit{hand over hand} stepping \cite{noauthor_drunk_nodate}. Dynein, however, is unique in its motion as it takes hand-over-hand steps but also \textit{inchworms} meaning that it often steps twice with the same foot. Furthermore, dynein even takes frequent backwards steps as shown in the stepping histogram in figure \ref{fig:weihong histogram} \cite{qiu_dynein_2012}. Collectively this behavior has earned dynein the title of \textit{drunken walker} which compactly summarizes its peculiar tendencies \cite{noauthor_drunk_nodate}. \\

@@

This unique stepping behavior makes dynein an interesting subject to study. There have been significant efforts aimed at characterizing the complex, multi-phase chemical cycle involved in the hydrolysis of ATP\cite{cianfrocco_mechanism_2015}. However, the dynamics of dynein's physical motion through space remains an open question. In this thesis, we examine a simplified physical model for the motion of dynein using Brownian dynamics. By simulating this model, we demonstrate it's ability to replicate of many observed properties of the protein. We will also use the model to comment a proposed mechanism for how dynein generates most of its movement, namely the powerstroke model.

>>>>

Cells are the fundamental unit for all life on Earth. In order
for them to perform their biological tasks, the complex structures they contain
must be able to replicate and move within their aqueous environment. This
transport problem  poses a unique challenge as diffusion alone is far to slow,
and random, a process to be able to produce movement on cellular timescales[].
Nature's solution to this problem is twofold. Within each cell there is a vast network of interweaving filaments called the \textit{cytoskeleton} that spans the gaps between the substructures of the cell. This cytoskeleton serves as a highway to a class of proteins called \textit{molecular motors} which can be distinguished into three types: kinesin, myosin, and dynein. Molecular motors attach to cellular cargo and drag it along the cytoskeleton by hydrolyzing adenosine triphosphate (ATP) in order to convert stored chemical energy into mechanical work\cite{alberts_molecular_2002}. Correct operation of these motors results in processes such as cell division, muscle contraction, and axonal transport; failure can lead to debilitating diseases such as ALS \cite{boillee_als:_2006}.

@@

Astonishingly, each of these motors is composed of two legs that combine to form a bipedal structure (see Figure \ref{fig:motor proteins}). that steps like a human as it moves along the cytoskeleton. Myosin and kinesin behave like a normal person; their steps alternate between opposing domains. This motion is called \textit{hand over hand} stepping \cite{noauthor_drunk_nodate}. Dynein, however, is unique in its motion as it takes hand-over-hand steps and also \textit{inchworms}. This alternative behavior is a scenario in which one motor consistently leads the opposite resulting in a shuffling motion. Furthermore, dynein even takes frequent backwards steps as shown in the stepping histogram in figure \ref{fig:weihong histogram} \cite{qiu_dynein_2012}. Collectively this behavior has earned dynein the title of \textit{drunken walker} which compactly summarizes its peculiar tendencies \cite{noauthor_drunk_nodate}. \\

@@

This unique stepping behavior makes dynein an interesting subject to study. There have been significant efforts aimed at characterizing the complex, multi-phase chemical cycle involved in the hydrolysis of ATP\cite{cianfrocco_mechanism_2015}. However, the dynamics of dynein's physical motion through space remains an open question as current optical techniques have difficulty resolving the individual motion of each dimer. In this thesis, we examine a simplified physical model for the motion of dynein using Brownian dynamics. By simulating this model, we demonstrate it's ability to replicate many of the observed properties of the protein. We will also use the model to comment a proposed mechanism for how dynein generates its movement, namely the powerstroke model.

### 2019-03-20 papers/waczak-thesis/thesis/chapters/theory.tex
big thesis update

	To do this we steer each domain towards an equilibrium configuration by a Hooke's law restorative torque with magnitude of the following form
	\begin{equation}
		\Gamma_i = -c_i\left(\varphi_i-\varphi_{i,\text{eq}}\right)
	\end{equation}
	Here the $c_i$ are spring constants (a free parameter of the model) and the set $\{\varphi_i\}$ are the angles measured between the $i-1$ and $i+1$ domains. Equally useful is the harmonic energy associated with each torque. These are given by
	\begin{equation}
		U_i = \frac{1}{2}c_i(\varphi_i-\varphi_{i,\text{eq}})^2
	\end{equation}

@@

	\section{Calculating Forces and Torques}
	Here I will discuss in detail how we calculate the forces and show how this is done in code

	\section{Equations of Motion}
	Now that we have specified the forces, the dynamics of each domain are determined by equation 4.5 so that in order to simulate the complicated structure we must solve a system of coupled first order differential equations. In doing so, it is convenient to introduce two states to distinguish between dynein's stepping and when both of its binding domains are attached to the microtubule. We shall refer to these configurations hereafter as the one bound state and the both bound state.

	In the one bound state, all but one of the domains are free to diffuse while the last is attached to the microtubule. In the both bound state, both binding domains are attached. This slight difference in geometry between the two scenarios leads to two sets of equations for motion that differ slightly. The following section details the derivation of these equations of motion.

@@

		The Brownian motion equation (3.5) also defines the velocity on domain.

@@

		\section{Determining constants of the model}
		The stalk and tail lengths $\ell_S, \ell_T$ as well as a set of equilibrium angles $\{\varphi_{i,\text{eq}}\}$ were determined by examining microscopy images of dynein in various configurations. 
		\begin{figure}[hbt!]
			\centering
			\includegraphics[width=0.3\columnwidth]{schematic-2-superimposed}
			\caption{Our model superimposed over microscopy image of Dynein \textbf{NOTE: add source} for determining lengths and equilibrium angles} 
			\label{fig:superimpmosed}
		\end{figure}
	\section{The Code}
		\subsection{The Random Force}

>>>>

	To do this we steer each domain towards an equilibrium configuration by a Hooke's law restorative torque with signed magnitude of the following form
	\begin{equation}
		\Gamma_i = -c_i\left(\varphi_i-\varphi_{i,\text{eq}}\right)
	\end{equation}
	Here the $c_i$ are spring constants (a free parameter of the model) and the set $\{\varphi_i\}$ are the angles measured between the $i-1$ and $i+1$ domains. Equally useful is the harmonic energy associated with each torque. These are given by
	\begin{equation}
		U_i = \frac{1}{2}c_i(\varphi_i-\varphi_{i,\text{eq}})^2
	\end{equation}

@@

	\section{Calculating Forces and Torques}
	However we decide to choose the forces on each domain, we would like them to satisfy  conservation angular momentum as this condition serves to keep the energy of the structure in check (add some stuff here) \todo[inline]{TODO: double check for other conditions on torques with David}

	Ultimately, we must specify the forces for three cases: a motor domain, and a tail domain, and the unbound binding domain. Having decided upon an algorithm for determining these forces, we could also calculate the force on a bound binding domain (attached to the microtubule) in order to estimate the maximum possible strain before unbinding occurs. This may prove interesting for future work but currently we do not care to calculate these forces as a bound binding domain is static and therefore, we do not need to evolve it's position.

	\begin{figure}[!hbt]
		\centering
		\includegraphics[angle =-90, width=0.65\columnwidth]{torque_forces.pdf}
		\caption{Toy model for three arbitrary domains. The force applied to adjacent domains due to the torque on the $i^{th}$ domain is shown. Forces are chosen to be perpendicular the inter-domain separation and in the direction that would move $\varphi_i$ towards its equilibrium configuration.}
		\label{fig:torque_force}
	\end{figure}

	Figure \ref{fig:torque_force} illustrates how the force on the motor domain due to the torque on the binding domain is chosen. In general, equation (3.10) suggests we should pick the magnitude of the force on the $i^{th}$ domain due to the torque on the $j=i\pm1$ domains to be
	\begin{equation}
		F_{ij} = \Gamma_j / \ell_{ij}
	\end{equation}
	where $\ell_{ij}$ is the distance between the $i$ and $j$ domains. We then make the choice to apply the force perpendicular to the inter-domain separation and in the direction of the equilibrium angle. Making this choice makes it clear that angular momentum will be conserved for the case where $\ell_{i\;i-1}=\ell_{i\;i+1}$. Finally to simplify computational complexity, we transform these two forces into forces on the $i^{th}$ domain by instead applying an equal and opposite force to each $\mathbf{F}_{ij}$ on the $i^{th}$ domain.

	With this algorithm, we can loop through each domain and calculate the torque-related forces on each domain. For binding domains, there is only one adjacent domain to consider. 
	\todo[inline]{TODO: add another figure or update above to indicate the equal-and-opposite forces that are actually kept. It may look nice to turn force arrows into dotted arrow and then add the force on the $i^{th}$ domain with a solid line}

	We are now armed with a way to calculate the forces on each domain due to the restorative torque. The only force which we are unable to calculate directly is the force due to tension. We will deal with these forces by applying the method Lagrange multipliers as we solve for the equations of motion.

	\section{Equations of Motion}
	The dynamics of each domain are determined by equation 3.5 so that in order to simulate the complicated structure we must solve a system of coupled first order differential equations. In doing so, it is convenient to introduce two states to distinguish between dynein's stepping and when both of its binding domains are attached to the microtubule. We shall refer to these configurations hereafter as the one bound state and the both bound state with the idea being that they directly correspond to the pre and post-stroke states defined by the mechanochemical cycle of \cite{cianfrocco_mechanism_2015}.

	In the one bound state, all but one of the domains are free to diffuse while the last is attached to the microtubule. In the both bound state, both binding domains are attached. This slight difference in geometry between the two scenarios leads to two independent equations motion motion that require careful consideration when transitioning between states. The following sections detail the derivation of these equations of motion for each configuration.

@@

		The Brownian motion equation (3.5) also defines the velocity on each domain allowing us to write
		\begin{align}
			\dot{x}_i &= \frac{1}{\gamma_i}\Big(F_{i,\Gamma_i}^x+ \lambda_{i,i-1}(x_i-x_{i-1})+\lambda_{i,i+1}(x_i-x_{i+1})-F_{i,\text{rand}}^x\Big)\\
			\dot{y}_i &= \frac{1}{\gamma_i}\Big(F_{i,\Gamma_i}^y+ \lambda_{i,i-1}(y_i-y_{i-1})+\lambda_{i,i+1}(y_i-y_{i+1})-F_{i,\text{rand}}^y\Big) \notag
		\end{align}
		where superscript $x$ and $y$ denote the force components.

		\todo[inline]{Double check this calculation with David to make sure I have accounted for the tension terms correctly. For some reason, Elliott uses x's in both the x and y equations... Is this a typo?}

		 At this point, comparing equations 3.13 with 3.14 reveals we have identified 8 nontrivial equations (x and y derivatives; ignoring bound binding domain 0). There are four unknown derivatives $\{\dot\theta_i\}$ and four unknown Lagrange multipliers $\{\lambda_{ij}\}$ which altogether form a linear system which we solve using Mathematica. For a detailed calculation see the appendices.

		 \todo[inline]{TODO: add appendices with full matrix equations}

@@

		\section{Determining constants of the model}
		The stalk and tail lengths $\ell_S, \ell_T$ as well as a set of equilibrium angles $\{\varphi_{i,\text{eq}}\}$ were determined by examining microscopy images of dynein in various configurations. 
		\begin{figure}[hbt!]
			\centering
			\includegraphics[width=0.3\columnwidth]{schematic-2-superimposed}
			\caption{Our model superimposed over microscopy image of Dynein \textbf{NOTE: add source} for determining lengths and equilibrium angles} 
			\label{fig:superimpmosed}
		\end{figure}

	\section{Binding and Unbinding}

	\todo[inline]{Add theoretical description of binding and unbinding probabilities and their relation to tension and torque on binding domain. Explain the use of our exponential factor and why it is reasonable. Discuss potential to add in exponential factor to binding rate.}

### 2019-03-20 papers/waczak-thesis/thesis/thesis.tex
thesis update
more thesis stuff
big thesis update

\documentclass[12pt]{report}
\usepackage{graphicx} % make it easy to load in images
\graphicspath{{./images/}} % where to find images
\usepackage[a4paper, margin=1in]{geometry} % change margins 
\usepackage{fancyhdr} % add a header / footer 
\pagestyle{fancy}

@@

\usepackage{caption}  % for creating simple subfigures
\usepackage{subcaption}
\title{
	{Simulating the Brownian Dynamics of a Two-Dimensional Model for the Dynein Motor Protein}\vspace{1.5em}\\
	\includegraphics[width=0.5\columnwidth]{OSU_logo.png}\vspace{0.5em}\\
	\large{Oregon State University}
}
\author{John Waczak}
\date{\today}

\begin{document}
	\maketitle

@@

	\chapter{Introduction}
	\input{./chapters/introduction}
	\chapter{Dynein}
	\input{./chapters/dynein}
	\chapter{Methods}
	\input{./chapters/methods}

\bibliography{references}
\bibliographystyle{acm}
\end{document}

>>>>

\documentclass[12pt]{report}
\usepackage{graphicx} % make it easy to load in images
\graphicspath{{./images/},{../../../plots/}} % where to find images
\usepackage[a4paper, margin=1in]{geometry} % change margins 
\usepackage{fancyhdr} % add a header / footer 
\pagestyle{fancy}

@@

\usepackage{caption}  % for creating simple subfigures
\usepackage{subcaption}

\usepackage{titlesec} % redefine headings to remove "Chapter X"

\usepackage[labelfont=bf, textfont=md]{caption}

\usepackage{setspace}

\usepackage{hyperref}

\usepackage[color=cyan]{todonotes}

\usepackage[labelfont=bf]{caption}

\title{
	{Simulating the Brownian Dynamics of a Two-dimensional Model for the Dynein Motor Protein}\vspace{1.5em}\\
	\includegraphics[width=0.5\columnwidth]{OSU_logo.png}\vspace{0.5em}\\
	\large{Oregon State University}
}
\author{John Waczak}
\date{\today}

\begin{document}
	\onehalfspacing

	\maketitle

@@

	\chapter{Introduction}
	\input{./chapters/introduction}
	\chapter{Dynein}
	\input{./chapters/dynein}
	\chapter{Theory}
	\input{./chapters/theory}
	\chapter{Methods}
	\input{./chapters/methods}
	\chapter{Results}
	\input{./chapters/results}
\bibliography{references}
\bibliographystyle{prl_like.bst}
\end{document}
